How is experience transformed into memory?

### 2018-09-02 paper/main.tex
minor polishing

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) is fundamentally different in at least three ways~\citep[also see][for a review]{KoriGold94}.  First, real world recall is continuous, rather than discrete.  Unlike in trial-based experiments, removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific words used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, precise recall is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the \textit{gist} or essence of an experience but forget (or neglect to recount) particular details.  Capturing the gist of what happened is typically the main ``point'' of recounting a memory to a listener whereas, depending on the circumstances, accurate recall of any specific detail may be irrelevant.  There is no analog of the gist of an experience in most traditional memory experiments; rather we tend to assess participants' abilities to recover specific details (e.g., the precise stimuli they encountered earlier in the experiment).

How might one go about formally characterizing the gist of an experience?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore the timecourse of how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g.\ the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  Although our experiences most often change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence-- i.e., their gists.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.

@@

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we analyzed the time-varying content of the video, to obtain similar estimates for participants' recalls we treated each (overlapping) 10 sentence ``window'' of their transcript as a ``document'' and then computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-sentences by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participant's recalls to events from the original video, despite that different participants may have used different language to describe the same event, and that those descriptions may not match the original annotations.  This is a huge benefit of projecting the video and recalls into a common ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

Although the example participant's recall topic proportions matrix has some visual similarity to the video topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as for the video (e.g., compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (e.g., most topics are active or inactive over contiguous blocks of time), the overall timecourses are not as cleanly delineated as the video topics are.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions (Fig.~\ref{fig:model}E).  As in the video correlation matrix (Fig.~\ref{fig:model}B), the example partipant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the video correlation matrix, we can use an HMM, along with the aforementioned event number optimization procedure (also see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

@@

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a common (shared) topic model and segmenting the content into events using HMMs.  Next we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented'' events from the original movie, and a sequence of (potentially differently ordered) ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[e.g., probability of first recall; Fig.~\listlearning A; ][]{WelcBurn24, PostPhil65, AtkiShif68}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[e.g., lag-conditional response probability; Fig.~\listlearning B; ][]{Kaha96}; and which events they were likely to remember overall~\citep[e.g., serial position recall analyses; Fig.~\listlearning C; ][]{Murd62a}.  In list-learning studies, this set of three analyses may be used to gain a nearly complete view into the sequences of recalls participants made~\citep[e.g., ][]{Kaha12}.  However, in naturalistic recall the analyses provide a wholly incomplete picture: they leave out any attempt to quantify participants' abilities to capture the \textit{content} of what occurred in the video (i.e., their only experimental instruction!).

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the video and recalls.}  All panels: the topic proportion matrices have been projected onto a common two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original video (see text), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  \textbf{C.} The recall topic trajectories taken by each individual participant (1--17).  The video's trajectory is shown in black for reference.  (Same format and coloring as Panel A.)}
\label{fig:trajectory}
\end{figure}

@@

Because our analysis framework projects the dynamic video content and participants' recalls onto a common topic space, and because the dimensions of that space are known (i.e., each topic dimension is a set of weights over words in the vocabulary; Fig.~\topics), we can examine the topic trajectories to understand which specific content was remembered well (or poorly).  For each video event, we can ask: for all participants who recalled that event, what was the average correlation (across participants) between the video event's topic vector and the recall event topic vectors from each participant?  This yields a single correlation coefficient for each video event, describing how closely participants' recalls of the event tended to reliably capture its content (Fig.~\ref{fig:topics}A).  (We also examined how different comparisons between each video event's topic vector and the corresponding recall event topic vectors related to hand-annotated characterizations of memory performance; see \textit{Supplemental Materials}).  Given this summary of which events were recalled reliably (or not), we next asked whether the better-remembered or worse-remembered events tended to reflect particular topics.  We computed a weighted average of the topic vectors for each video event, where the weights reflected how reliably each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by better-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Events that reflected topics weighting heavily on characters like ``Sherlock'' and ``John'' (i.e., the main characters) and locations like ``221b Baker Street'' (i.e., a major recurring location; the address of the flat that Sherlock and John share) were best remembered.  An analogous analysis revealed which themes were poorly remembered; here in computing the weighted average over events' topic vectors we weighted each event in \textit{inverse} proportion to how well it was remembered (Fig.~\ref{fig:topics}B, red box).  This revealed that events with relatively minor characters such as ``Mike,'' ``Jeffrey,'' and ``Molly,'' as well as less-integral plot locations (e.g., ``hospital'' and ``office'') were least well-remembered.  This suggests that what is retained in memory are the major plot elements (i.e., the overall ``gist'' of what happened), whereas the more minor details are prone to pruning.

>>>>

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) is fundamentally different in at least three ways~\citep[also see][for a review]{KoriGold94}.  First, real world recall is continuous, rather than discrete.  Unlike in trial-based experiments, removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific words used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, counting the number or proportion of precise recalls is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the \textit{gist} or essence of an experience but forget (or neglect to recount) particular details.  Capturing the gist of what happened is typically the main ``point'' of recounting a memory to a listener whereas, depending on the circumstances, accurate recall of any specific detail may be irrelevant.  There is no analog of the gist of an experience in most traditional memory experiments; rather we tend to assess participants' abilities to recover specific details (e.g., computing the proportion of specific stimuli they remember, which presentation positions the remembered stimuli came from, etc.).

How might one go about formally characterizing the gist of an experience, or whether that gist has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore the timecourse of how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g.\ the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  Although our experiences most often change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence-- i.e., their gists.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.

@@

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we analyzed the time-varying content of the video, to obtain similar estimates for participants' recalls we treated each (overlapping) 10 sentence ``window'' of their transcript as a ``document'' and then computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-sentences by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participant's recalls to events from the original video, despite that different participants may have used different language to describe the same event, and that those descriptions may not match the original annotations.  This is a huge benefit of projecting the video and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

Although the example participant's recall topic proportions matrix has some visual similarity to the video topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as for the video (e.g., compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (e.g., most topics are active or inactive over contiguous blocks of time), the overall timecourses are not as cleanly delineated as the video topics are.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions (Fig.~\ref{fig:model}E).  As in the video correlation matrix (Fig.~\ref{fig:model}B), the example partipant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the video correlation matrix, we can use an HMM, along with the aforementioned number-of-events optimization procedure (also see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

@@

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events using HMMs.  Next we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented'' events from the original movie, and a sequence of (potentially differently ordered) ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[e.g., probability of first recall; Fig.~\listlearning A; ][]{WelcBurn24, PostPhil65, AtkiShif68}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[e.g., lag-conditional response probability; Fig.~\listlearning B; ][]{Kaha96}; and which events they were likely to remember overall~\citep[e.g., serial position recall analyses; Fig.~\listlearning C; ][]{Murd62a}.  In list-learning studies, this set of three analyses may be used to gain a nearly complete view into the sequences of recalls participants made~\citep[e.g., ][]{Kaha12}.  However, in naturalistic recall the analyses provide a wholly incomplete picture: they leave out any attempt to quantify participants' abilities to capture the \textit{content} of what occurred in the video (i.e., their only experimental instruction!).

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the video and recalls.}  All panels: the topic proportion matrices have been projected onto a shared two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original video (see text), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  \textbf{C.} The recall topic trajectories taken by each individual participant (1--17).  The video's trajectory is shown in black for reference.  (Same format and coloring as Panel A.)}
\label{fig:trajectory}
\end{figure}

@@

Because our analysis framework projects the dynamic video content and participants' recalls onto a shared topic space, and because the dimensions of that space are known (i.e., each topic dimension is a set of weights over words in the vocabulary; Fig.~\topics), we can examine the topic trajectories to understand which specific content was remembered well (or poorly).  For each video event, we can ask: for all participants who recalled that event, what was the average correlation (across participants) between the video event's topic vector and the recall event topic vectors from each participant?  This yields a single correlation coefficient for each video event, describing how closely participants' recalls of the event tended to reliably capture its content (Fig.~\ref{fig:topics}A).  (We also examined how different comparisons between each video event's topic vector and the corresponding recall event topic vectors related to hand-annotated characterizations of memory performance; see \textit{Supplemental Materials}).  Given this summary of which events were recalled reliably (or not), we next asked whether the better-remembered or worse-remembered events tended to reflect particular topics.  We computed a weighted average of the topic vectors for each video event, where the weights reflected how reliably each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by better-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Events that reflected topics weighting heavily on characters like ``Sherlock'' and ``John'' (i.e., the main characters) and locations like ``221b Baker Street'' (i.e., a major recurring location; the address of the flat that Sherlock and John share) were best remembered.  An analogous analysis revealed which themes were poorly remembered; here in computing the weighted average over events' topic vectors we weighted each event in \textit{inverse} proportion to how well it was remembered (Fig.~\ref{fig:topics}B, red box).  This revealed that events with relatively minor characters such as ``Mike,'' ``Jeffrey,'' and ``Molly,'' as well as less-integral plot locations (e.g., ``hospital'' and ``office'') were least well-remembered.  This suggests that what is retained in memory are the major plot elements (i.e., the overall ``gist'' of what happened), whereas the more minor details are prone to pruning.
